Fisher’s Method

### 2018-07-19 Fishers_method/Fishers_method_short.Rmd
added Rrita's name :)
changed subscripts

---
title: "Fisher’s Method"
author: "Katie Fitzgerald"
date: "06/20/2018"
output: pdf_document
---

@@

Among the 100 replication studies conducted by the Open Science Collaboration (OSC), 64 found null effects, defined as having a p-value greater than 0.05. The OSC applied Fisher's method to this set of non-significant p-values to test the null hypothesis that a true zero effect held for each study and that there were no false negatives among them. Therefore, their hypothesis for Fisher's test could be formalized as follows, where $\theta_i$ is the treatment effect for study $i$:
\begin{center}
$H_0$: $\theta_1=\theta_2=...=\theta_{64}=0$

$H_a$: at least one $\theta_i\neq 0,\quad i=1,...,64$
\end{center}

Because the OSC conducted Fisher's test only among the 64 replication studies that did not have significant results, and thus $p_i \geq 0.05$ for all $i$, they used the following transformation of Fisher's test statistic  $$X^2=-2\sum_{i=1}^{k} ln(p_i^*)=-2\sum_{i=1}^{k} ln(\frac{p_i-0.05}{0.95}),$$ where under $H_0$, the $p_i^*$'s follow a uniform distribution on [0,1] and thus $X^2 \sim \chi^2_{2k}$, where $k$ is the number of studies (in this case $k=64$). Low p-values give a larger test statistic, leading to a rejection of $H_0$.

@@

Even though the OSC was able to reject their null hypothesis ($X^2=155.83, p=0.048$), we think that in general Fisher's method is underpowered to answer the question at hand. We hypothesize that it requires many and possibly large non-null effects in order to skew the distribution of p-values enough to reject Fisher's null hypothesis. In the scenario where a researcher is combining the results of $k$ studies of the same treatment to test if an overall treatment effect $\Theta=0$, this type of conservative test may be appropriate. In the OSC scenario, however, since the 64 studies are not testing the same treatment effect, and one $\theta_i$ has no bearing on the 63 other $\theta_i's$, the presumed goal would be to detect if there are *any* false negatives among the replicate studies.

@@

For simplicity of interpretation but without loss of generality, we will work with the standardized scale of Cohen's d, defined as $\delta_i=\frac{\theta_i}{\sigma_i}$, where $\theta_i$ is the mean difference between the treatment and control groups in study $i$, and $\sigma_i$ is the known and equal variance among the treatment and control populations. Assuming equal sample sizes in the treatment and control groups within study $i$ (that is, let $n_i^t=n_i^c=n_i$), $\delta_i$ is estimated by $d_i\sim N(\delta_i,\frac{2}{n_i})$.[^2] Under this framework, Fisher's method can be represented in Cohen's d as testing the hypotheses

@@

$H_a$: at least one $\delta_i\neq 0,\quad i=1,...,64,$
\end{center} 
and the 64 p-values to be summed in Fisher's test statistic can be calculated as $p_i=2(1-\Phi(\frac{|d_i|}{\sqrt{2/n_i}})).$

We first consider the power of Fisher's method when $n_i=76/2=38$ for all $i$, because 76 is the median sample size in the OSC dataset, and we are assuming equal sample sizes in the treatment and control groups. The results of the power simulations for median sample size are given in Table 1.[^3] Note that 0.2, 0.5, and 0.8 correspond to small, medium, and large effects sizes on the scale of Cohen's d. As shown in the last column of Table 1, we find that there needs to be 10 large effects in order to achieve close to the standard 80% power (power=0.7743). If there is only one large effect, Fisher's test only has approximately 8% power to reject. Even when all 64 studies have small effects, Fisher's method only has 70% power to reject.

@@

Table: Power of Fisher's method given median sample size $(n_i=38)$ for varying $\delta$ and true # of non-null effects
```{r,echo=FALSE}
library(knitr)
table <- readRDS("./power sims table with changing n.RDS")
table_median<-table[,1:4]
#colnames(table_median)[2]<
kable(table_median,format="markdown",digits=4,align=c('l','r','r','r'),col.names=c("# of non-null effects","$\\delta=0.2$","$\\delta=0.5$","$\\delta=0.8$"), caption="Power to reject H0 with median sample size (n=38)")

@@

Table: Power of Fisher's method given large sample sizes for varying $\delta$ and true # of non-null effects (i.e. "Best-case scenario")
```{r, echo=FALSE}
table <- readRDS("./power_large_n.RDS")
kable(table,format="markdown",digits=4,align=c('l','r','r','r'),col.names=c("# of non-null effects","$\\delta=0.2$","$\\delta=0.5$","$\\delta=0.8$"),caption="Power to reject H0 with large n")
```
[^4]:Note that because we are working with a set of replicate studies which found a p-value greater than 0.05, pairing large effects with very large sample sizes is not realistic, and therefore we begin the simulations with the largest $n_i$ among the studies for which the power is at most 99.99% to detect the given $\delta_i$. The largest sample sizes used for $\delta_i=0.2; 0.5$; and $0.8$ were $n_i=745; 159;$ and $100$ respectively. See Appendix A.3 for further discussion.

@@

Assume each study $i$ is testing the presence of some treatment effect $\theta_i$, where $i=1,...,k$, and $k$ is the number of studies. Assume equal sample sizes for the treatment and control groups ($n_{ti}=n_{ci}=n_i$). Let $Y_{ij}^t$ and $Y_{ij}^c$ be the observations from the treatment and control groups respectively ($i=1,...,k$, and $j=1,...n_i$), and assume $Y_{ij}^t\sim N(\mu_i +\theta_i, \sigma_i^2)$ and $Y_{ij}^c\sim N(\mu_i, \sigma_i^2)$, where the $\sigma_i^2$'s are known. Note this is not an unreasonbale assumption since this framework is required for both the t-test and the ANOVA test, and the majority of the tests in the OSC subset of 64 replicate studies are of these two types (89%).

Note then, $\theta_i$ is the difference in means between the treatment and control groups, and its estimate $T_i=\overline{Y_{i.}^t}-\overline{Y_{i.}^c}$ has variance $v_i=\frac{\sigma^2}{n_{ci}} +\frac{\sigma^2}{n_{ti}}=\frac{2\sigma^2}{n_i}.$ For simplicity of interpretation but without loss of generality, we work with the standardized scale of Cohen's d, defined as $\delta_i=\frac{\theta_i}{\sigma}$. Note then that $\delta_i$ is estimated by $d_i=\frac{T_i}{\sigma}=\frac{T_i}{\sqrt{\frac{v_i}{2/n_i}}}=\frac{T_i}{\sqrt{v_i}}\sqrt{{\frac{2}{n_i}}}.$ Since under $H_0$, $T_i\sim N(\theta_i,v_i) \Rightarrow \frac{T_i}{\sqrt{v_i}} \sim N(\frac{\theta_i}{\sqrt{v_i}},1)$ and $\sqrt{\frac{2}{n_i}}$ is a constant, then we have  $Var(d_i)=Var(\frac{T_i}{\sqrt{v_i}}\sqrt{{\frac{2}{n_i}}})=\frac{2}{n_i}$. Therefore, $d_i\sim N(\delta_i,\frac{2}{n_i})$.

###A.2 Power simulation logic and code
Let there be $m$ false negatives (i.e. $m$ true non-zero effects) among the 64 studies, $m=1,...,64$. Therefore we must draw $m$ p-values from a distribution consistent with the alternative hypothesis. That is, we draw a random variable $d_i$ from a $N(\delta_i,\frac{2}{n_i})$ distribution, where $\delta_i\neq0$ and compute its p-value. We continue drawing $d_i's$ until we obtain $m$ p-values greater than 0.05 (due to the OSC restriction of only considering replicate studies with non-significant results). We will draw the remaining $64-m$ p-values from a U[0.05,1] distribution and then calculate Fisher's test statistic $X^2=-2\sum_{i=0}^{k} ln(\frac{p_i-0.05}{0.95})$. We run this procedure N times and calculate the simulated power of Fisher's method under these conditions to be $\sum_{l=0}^{N}I_{\{X^2_l>155.4047\}}/N,$ where $I$ is the indicator function and $\chi^2_{128}=155.4047$ is the critical value for Fisher's test with $k=64$ studies. We let N=100,000. The code is given below.

@@

As noted in footnote 4 of the text, the largest sample sizes were dropped out of necessity in the "best-case scenario" power simulations presented in Table 2. For example, the largest study had $n_i=384351.5$, which is powered at 100% to detect even a small $\delta=0.2$. Therefore, it would have been impossible for this study to result in a p-value greater than 0.05 if there was a true non-null effect. Table 3 presents the power of the OSC replicate studies to detect a given $\delta$ using the true sample sizes for the 64 studies. For each $\delta$, we began the power simulations with the largest $n$ for which the power was at most 0.9999. The largest sample sizes used for $\delta_i=0.2; 0.5$; and $0.8$ therefore were $n_i=745; 159;$ and $100$ respectively. Table 4 presents the sample size vectors used for the "best-case scenario" power simulations.

>>>>

---
title: "Fisher’s Method"
author: "Katie Fitzgerald and Rrita Zejnullahi"
date: "06/20/2018"
output: pdf_document
---

@@

Among the 100 replication studies conducted by the Open Science Collaboration (OSC), 64 found null effects, defined as having a p-value greater than 0.05. The OSC applied Fisher's method to this set of non-significant p-values to test the null hypothesis that a true zero effect held for each study and that there were no false negatives among them. Therefore, their hypothesis for Fisher's test could be formalized as follows, where $\theta_j$ is the treatment effect for study $j$:
\begin{center}
$H_0$: $\theta_1=\theta_2=...=\theta_{64}=0$

$H_a$: at least one $\theta_j\neq 0,\quad j=1,...,64$
\end{center}

Because the OSC conducted Fisher's test only among the 64 replication studies that did not have significant results, and thus $p_j \geq 0.05$ for all $j$, they used the following transformation of Fisher's test statistic  $$X^2=-2\sum_{j=1}^{k} ln(p_j^*)=-2\sum_{j=1}^{k} ln(\frac{p_j-0.05}{0.95}),$$ where under $H_0$, the $p_j^*$'s follow a uniform distribution on [0,1] and thus $X^2 \sim \chi^2_{2k}$, where $k$ is the number of studies (in this case $k=64$). Low p-values give a larger test statistic, leading to a rejection of $H_0$.

@@

Even though the OSC was able to reject their null hypothesis ($X^2=155.83, p=0.048$), we think that in general Fisher's method is underpowered to answer the question at hand. We hypothesize that it requires many and possibly large non-null effects in order to skew the distribution of p-values enough to reject Fisher's null hypothesis. In the scenario where a researcher is combining the results of $k$ studies of the same treatment to test if an overall treatment effect $\Theta=0$, this type of conservative test may be appropriate. In the OSC scenario, however, since the 64 studies are not testing the same treatment effect, and one $\theta_j$ has no bearing on the 63 other $\theta_j's$, the presumed goal would be to detect if there are *any* false negatives among the replicate studies.

@@

For simplicity of interpretation but without loss of generality, we will work with the standardized scale of Cohen's d, defined as $\delta_j=\frac{\theta_j}{\sigma_j}$, where $\theta_j$ is the mean difference between the treatment and control groups in study $j$, and $\sigma_j$ is the known and equal variance among the treatment and control populations. Assuming equal sample sizes in the treatment and control groups within study $j$ (that is, let $n_j^t=n_j^c=n_j$), $\delta_j$ is estimated by $d_j\sim N(\delta_j,\frac{2}{n_j})$.[^2] Under this framework, Fisher's method can be represented in Cohen's d as testing the hypotheses

@@

$H_a$: at least one $\delta_j\neq 0,\quad j=1,...,64,$
\end{center} 
and the 64 p-values to be summed in Fisher's test statistic can be calculated as $p_j=2(1-\Phi(\frac{|d_j|}{\sqrt{2/n_j}})).$

We first consider the power of Fisher's method when $n_j=76/2=38$ for all $j$, because 76 is the median sample size in the OSC dataset, and we are assuming equal sample sizes in the treatment and control groups. The results of the power simulations for median sample size are given in Table 1.[^3] Note that 0.2, 0.5, and 0.8 correspond to small, medium, and large effects sizes on the scale of Cohen's d. As shown in the last column of Table 1, we find that there needs to be 10 large effects in order to achieve close to the standard 80% power (power=0.7743). If there is only one large effect, Fisher's test only has approximately 8% power to reject. Even when all 64 studies have small effects, Fisher's method only has 70% power to reject.

@@

Table: Power of Fisher's method given median sample size $(n_j=38)$ for varying $\delta$ and true # of non-null effects
```{r,echo=FALSE}
library(knitr)
table <- readRDS("./power sims table with changing n.RDS")
table_median<-table[,1:4]
#colnames(table_median)[2]<
kable(table_median,format="markdown",digits=4,align=c('l','r','r','r'),col.names=c("# of non-null effects","$\\delta=0.2$","$\\delta=0.5$","$\\delta=0.8$"), caption="Power to reject H0 with median sample size (n=38)")

@@

Table: Power of Fisher's method given large sample sizes for varying $\delta$ and true # of non-null effects (i.e. "Best-case scenario")
```{r, echo=FALSE}
table <- readRDS("./power_large_n.RDS")
kable(table,format="markdown",digits=4,align=c('l','r','r','r'),col.names=c("# of non-null effects","$\\delta=0.2$","$\\delta=0.5$","$\\delta=0.8$"),caption="Power to reject H0 with large n")
```
[^4]:Note that because we are working with a set of replicate studies which found a p-value greater than 0.05, pairing large effects with very large sample sizes is not realistic, and therefore we begin the simulations with the largest $n_j$ among the studies for which the power is at most 99.99% to detect the given $\delta_j$. The largest sample sizes used for $\delta_j=0.2; 0.5$; and $0.8$ were $n_j=745; 159;$ and $100$ respectively. See Appendix A.3 for further discussion.

@@

Assume each study $j$ is testing the presence of some treatment effect $\theta_j$, where $j=1,...,k$, and $k$ is the number of studies. Assume equal sample sizes for the treatment and control groups ($n_{tj}=n_{cj}=n_j$). Let $Y_{ij}^t$ and $Y_{ij}^c$ be the observations from the treatment and control groups respectively ($i=1,...,k$, and $j=1,...n_i$), and assume $Y_{jl}^t\sim N(\mu_j +\theta_j, \sigma_j^2)$ and $Y_{jl}^c\sim N(\mu_j, \sigma_j^2)$, where the $\sigma_j^2$'s are known. Note this is not an unreasonbale assumption since this framework is required for both the t-test and the ANOVA test, and the majority of the tests in the OSC subset of 64 replicate studies are of these two types (89%).

Note then, $\theta_j$ is the difference in means between the treatment and control groups, and its estimate $T_j=\overline{Y_{j.}^t}-\overline{Y_{j.}^c}$ has variance $v_j=\frac{\sigma^2}{n_{cj}} +\frac{\sigma^2}{n_{tj}}=\frac{2\sigma^2}{n_j}.$ For simplicity of interpretation but without loss of generality, we work with the standardized scale of Cohen's d, defined as $\delta_j=\frac{\theta_j}{\sigma}$. Note then that $\delta_j$ is estimated by $d_j=\frac{T_j}{\sigma}=\frac{T_j}{\sqrt{\frac{v_j}{2/n_j}}}=\frac{T_j}{\sqrt{v_j}}\sqrt{{\frac{2}{n_j}}}.$ Since under $H_0$, $T_j\sim N(\theta_j,v_j) \Rightarrow \frac{T_j}{\sqrt{v_j}} \sim N(\frac{\theta_j}{\sqrt{v_j}},1)$ and $\sqrt{\frac{2}{n_j}}$ is a constant, then we have  $Var(d_j)=Var(\frac{T_j}{\sqrt{v_j}}\sqrt{{\frac{2}{n_j}}})=\frac{2}{n_j}$. Therefore, $d_j\sim N(\delta_j,\frac{2}{n_j})$.

###A.2 Power simulation logic and code
Let there be $m$ false negatives (i.e. $m$ true non-zero effects) among the 64 studies, $m=1,...,64$. Therefore we must draw $m$ p-values from a distribution consistent with the alternative hypothesis. That is, we draw a random variable $d_j$ from a $N(\delta_j,\frac{2}{n_j})$ distribution, where $\delta_j\neq0$ and compute its p-value. We continue drawing $d_j's$ until we obtain $m$ p-values greater than 0.05 (due to the OSC restriction of only considering replicate studies with non-significant results). We will draw the remaining $64-m$ p-values from a U[0.05,1] distribution and then calculate Fisher's test statistic $X^2=-2\sum_{j=1}^{k} ln(\frac{p_j-0.05}{0.95})$. We run this procedure N times and calculate the simulated power of Fisher's method under these conditions to be $\sum_{q=1}^{N}I_{\{X^2_q>155.4047\}}/N,$ where $I$ is the indicator function and $\chi^2_{128}=155.4047$ is the critical value for Fisher's test with $k=64$ studies. We let N=100,000. The code is given below.

@@

As noted in footnote 4 of the text, the largest sample sizes were dropped out of necessity in the "best-case scenario" power simulations presented in Table 2. For example, the largest study had $n_j=384351.5$, which is powered at 100% to detect even a small $\delta=0.2$. Therefore, it would have been impossible for this study to result in a p-value greater than 0.05 if there was a true non-null effect. Table 3 presents the power of the OSC replicate studies to detect a given $\delta$ using the true sample sizes for the 64 studies. For each $\delta$, we began the power simulations with the largest $n$ for which the power was at most 0.9999. The largest sample sizes used for $\delta_j=0.2; 0.5$; and $0.8$ therefore were $n_j=745; 159;$ and $100$ respectively. Table 4 presents the sample size vectors used for the "best-case scenario" power simulations.
